\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-01-30 move_schedules.tex
Added performance images

%% TODO: elaborate on monotonicity, describe segment tree
The simplicity of both the Spearman distance and the types of permutations we consider here enables certain computational advantages. Suppose we begin with an array $\mathcal{A}$ of size $m$ which provides $O(1)$ access and modification, initialized with the (signed) \emph{displacement} of every element in $p$ to its corresponding position in $q$. Since the Spearman distance is simply the sum of the absolute value of these displacements, at any point during during the execution of Algorithm~\ref{alg:sorting} we may obtain $F(\bar{p}, q)$ simply by having access to the sum of every entry in $\mathcal{A}$. There are two degrees of freedom in   Algorithm~\ref{alg:sorting}: the first is in the choice of $\sigma \in \mathcal{D}$, and the second is in choosing $j$. If we fix a heuristic for the latter, then we have a set of possible valid moves $S_\mathcal{D}$ induced by each $\sigma \in \mathcal{D}$ to choose from. Observe that each permutation $s_\sigma \in S_\mathcal{D}$ changes the displacement of every symbol in at most three different ways: 
\[
\mathcal{A}(s_\sigma \circ p) = 
\begin{cases} 
	 \mathcal{A}(\sigma) \pm \lvert i - j \rvert & p^{-1}(\sigma) = i \\
	 \mathcal{A}(\sigma) \pm 1 & i < p^{-1}(\sigma) \leq j \\
	 \mathcal{A}(\sigma) & \text{ otherwise }
\end{cases}
\]
That is, if $s_\sigma$ represents a move permutation moving $\sigma$ from position $i$ to position $j$, then relative to the previous permutation $p$, the displacement of $\sigma$ in $s_\sigma \circ p$ changes by $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. 
All other symbol displacements are unchanged. Thus, if we replace $\mathcal{A}$ with a data structure  supporting $O(\log m)$ access time to aggregate information (i.e. the sum) and $O(\log m)$ modification time ability on $\lvert i - j \vert$ entries (simultaneously), we could greedily choose the next permutation $s$ to maximize:
\begin{equation}\label{eq:greedy_step}
	s_{\text{greedy}} = \argmin_{\text{ valid } s \in S_\mathcal{D}} F(s\circ p, q)
\end{equation}  
in $O(d\log m)$ time. 
The former problem reduces to the problem of efficiently calculating \emph{prefix sums}, which is easily solved. 
It is not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$.  
% Can actually only achieve O(d^2 log m) it seems 
It is known that one can apply constant-factor updates to multiple values in a prefix sum in $O(\log m)$ time by any prefix-sum data structure that also supports \emph{range updates}---e.g. an \emph{implicit treap}. 
Since single element modifications, removals, and insertions can all be achieved in $O(\log m)$ expected time with such a data structure, we conclude that a greedy solution to equation~\eqref{eq:greedy_step} can be computed in $O(d^2 \log m)$ time. 
%Since single element prefix sum modifications can also be achieved $O(\log m)$ time with these data structures, the time to compute $F(\hat{s}_i, \hat{s}_{i+1})$ is bounded above by $O(\log m)$, 
%thus the greedy solution to equation~\ref{eq:schedule_opt_succ} can be computed in $O(d^2 \log m)$ time.

@@

% TODO: Move computational results to subsection, add applications
% \subsection{Empirical results}\label{sec:results}
% \section{Applications and Discussion}\label{sec:applications}
\section{Applications and Experiments}\label{sec:results}
\subsection*{Crocker stacks}
There are many challenges to characterizing topological behavior in dynamic settings. One approach is to trace out the curves constituting a continuous family of persistence diagrams in $\mathbb{R}^3$---the vineyards approach---however this visualization can be cumbersome to work with as there are potentially many such vines tangled together, making topological critical events with low persistence difficult to detect. Moreover, the vineyards visualization does not admit a natural simplification utilizing the stability properties of persistence, as individual vines are not stable: if two vines move near each other and then pull apart without touching, then a pairing in their corresponding persistence diagrams may cross under a small perturbation, signaling the presence of an erroneous topological critical event~\cite{topaz2015topological, xian2020capturing}.

>>>>

%% TODO: elaborate on monotonicity, describe segment tree
We propose a heuristic strategy to minimize~\eqref{eq:sorting_additive} which we observed is both efficient to compute quite effective in practice.
The heuristic is inspired by the simplicity of computing the Spearman distance between cyclic permutations at the schedule construction phase. Suppose we begin with an array $\mathcal{A}$ of size $m$ which provides $O(1)$ access and modification, initialized with the (signed) \emph{displacement} of every element in $p$ to its corresponding position in $q$. Since the Spearman distance is simply the sum of the absolute value of these displacements, at any point during during the execution of Algorithm~\ref{alg:sorting} we may obtain $F(\bar{p}, q)$ simply by having access to the sum of every entry in $\mathcal{A}$. There are two degrees of freedom in   Algorithm~\ref{alg:sorting}: the first is in the choice of $\sigma \in \mathcal{D}$, and the second is in choosing $j$. If we fix a heuristic for the latter, then we have a set of possible valid moves $S_\mathcal{D}$ induced by each $\sigma \in \mathcal{D}$ to choose from. Observe that each permutation $s_\sigma \in S_\mathcal{D}$ changes the displacement of every symbol in at most three different ways: 
\[
\mathcal{A}(s_\sigma \circ p) = 
\begin{cases} 
	 \mathcal{A}(\sigma) \pm \lvert i - j \rvert & p^{-1}(\sigma) = i \\
	 \mathcal{A}(\sigma) \pm 1 & i < p^{-1}(\sigma) \leq j \\
	 \mathcal{A}(\sigma) & \text{ otherwise }
\end{cases}
\]
That is, if $s_\sigma$ represents a move permutation moving $\sigma$ from position $i$ to position $j$, then relative to the previous permutation $p$, the displacement of $\sigma$ in $s_\sigma \circ p$ changes by $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. 
All other symbol displacements are unchanged. Thus, if we replace $\mathcal{A}$ with a data structure  supporting $O(\log m)$ access time to aggregate information (i.e. the sum) and $O(\log m)$ modification time ability on $\lvert i - j \vert$ entries (simultaneously), we could greedily choose the next permutation $s$ to maximize:
\begin{equation}\label{eq:greedy_step}
	s_{\text{greedy}} = \argmin_{\text{ valid } s \in \Sigma} F(s\circ p, \iota)
\end{equation}  
in $O(d\log m)$ time. 
The former problem reduces to the problem of efficiently calculating \emph{prefix sums}, which is easily solved. 
It is not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$.  
% Can actually only achieve O(d^2 log m) it seems 
It is known that one can apply constant-factor updates to multiple values in a prefix sum in $O(\log m)$ time by any prefix-sum data structure that also supports \emph{range updates}---e.g. an \emph{implicit treap}. 
Since single element modifications, removals, and insertions can all be achieved in $O(\log m)$ expected time with such a data structure, we conclude that a greedy solution to equation~\eqref{eq:greedy_step} can be computed in $O(d^2 \log m)$ time. 
%Since single element prefix sum modifications can also be achieved $O(\log m)$ time with these data structures, the time to compute $F(\hat{s}_i, \hat{s}_{i+1})$ is bounded above by $O(\log m)$, 
%thus the greedy solution to equation~\ref{eq:schedule_opt_succ} can be computed in $O(d^2 \log m)$ time.

@@

% TODO: Move computational results to subsection, add applications
% \subsection{Empirical results}\label{sec:results}
% \section{Applications and Discussion}\label{sec:applications}
\section{Applications and Experiments}\label{sec:results}
\subsection*{Movies}
Move data description
\begin{figure}
	\centering
	\includegraphics[width=0.48\textwidth]{varying_d.pdf}
	\includegraphics[width=0.48\textwidth]{varying_nd.pdf}
	\caption{Performance}
\end{figure}

\subsection*{Crocker stacks}
There are many challenges to characterizing topological behavior in dynamic settings. One approach is to trace out the curves constituting a continuous family of persistence diagrams in $\mathbb{R}^3$---the vineyards approach---however this visualization can be cumbersome to work with as there are potentially many such vines tangled together, making topological critical events with low persistence difficult to detect. Moreover, the vineyards visualization does not admit a natural simplification utilizing the stability properties of persistence, as individual vines are not stable: if two vines move near each other and then pull apart without touching, then a pairing in their corresponding persistence diagrams may cross under a small perturbation, signaling the presence of an erroneous topological critical event~\cite{topaz2015topological, xian2020capturing}.
